Network based multifactorial modelling microRNA-target interaction.

### 2019-09-12 article.Rmd
news to article
edited with selcen

Competing endogenous RNA (ceRNA) regulations and crosstalk between various types of non-coding RNA in human is remarkable in means of miRNA regulation. Many studies have pointed out that an alteration in miRNA:target interaction can result in unexpected changes due to indirect and complex interactions. In this paper, we defined a new network-based model that handles miRNA:ceRNA interactions with expression values. Our model is able to handle miRNA interaction factors such as seed type, binding energy, if provided. Our approach is able to reveal that a perturbation in an element of network affects whole competing elements differently and cooperative efficiencies of miRNAs on common targets could be calculated. Our findings emphasized importance of miRNA:target ratios being crucial, as reported by previous studies. We have showed that the competing elements which have the same or close expression values may not be affected equally from the perturbation because of repression functionality depended on interaction factors of miRNA target pairs (answer: Hocam etkileşim faktörlerinden dolayı ekspresyonları yakın olsa bile baskılanma aktitesinin aynı sergilenmeyeceğini ifade etmeye çalıştım. Bu şekilde anlaşılıyor mu?). We applied the model to real sample consisting of breast cancer gene and miRNA expression dataset and experimental miRNA:target interaction dataset all generated via high throughput sequencing methods. A gene over-expressed in tumor tissue, namely *ABCC1*, is used as perturbing element. We have observed that change in expression level of single gene in miRNA:target network is sufficient to perturb regulations in whole network, due to unforeseen and unpredicted regulation which are only visible when considered in network context. Therefore, this model helps unveiling the crosstalk between elements in miRNA:target network where abundance of target and sponge effect are taken into account. The model is scalable and can be plugged in with emerging miRNA effectors such as circRNAs. The model is available as R package [ceRNAnetsim](https://github.com/selcenari/ceRNAnetsim).

@@

MicroRNAs (miRNAs) are a family of short non-coding RNAs which are key regulator of gene expression through various post-transcriptional mechanisms. Although the mechanisms by which miRNA represses are not fully understood, miRNAs predominantly repress their targets. Repressive activities of miRNAs vary depending on many factors that are significant to miRNA:target interactions. These factors include miRNA:target binding energy, binding location in target sequence, base pairing types between miRNA and target, abundance of miRNAs and targets [@grimson_microrna_2007]. Binding energies of miRNA:target complexes vary based on nucleotide context and determine folding stability of complex [@cao_predicting_2012]. It has been demonstrated that the binding energy between miRNA and target indicates stability or affinity of complex [@helwak_mapping_2013] and does not directly determine repressive activity of miRNA [@cao_predicting_2012]. Early studies have argued that 2-8 nt sequence, seed, located in miRNA 5’end bind to specific sequence located in 3’UTR of its target [@bartel_micrornas_2004; @lewis_conserved_2005]. In recent studies, it has been shown that miRNAs can interact with targets via sequences located in regions such as 5’UTR or CDS [@hausser_analysis_2013; @helwak_mapping_2013; @moore_mirnatarget_2015]. These studies also showed that binding location could indicate functionality of miRNA:target interaction or be effective on abundance of targets(TOFIX abundance etkisi derken, bazen azaltabilir, bazen arttırabilir, manasında mı? answer: evet hocam anlaşılmıyorsa değiştirebilirim). It has been shown that miRNAs exhibit repressive activity via, 6-8 nt long sequence that is perfectly complementary with targets, seed region at the 5’ end of miRNAs [@bartel_micrornas:_2009; @grimson_microrna_2007]. On the other hand, some researchers have reported that seed sequence of miRNA can have mismatches or bulged/wobble nucleotides [@chi2012alternative] and may locate in region other than 5'end of miRNAs [@hafner_transcriptome-wide_2010; @helwak_mapping_2013]. On top of all these factors, abundance of miRNAs and targets and miRNA:target ratio in cells predominantly affect efficiency of miRNA:target interaction [@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014].

@@

![Schematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: gray triangle) repress targets (G: circle coconut) according to proportion of their targets' expression. b) Two fold increase in transcript level of Gene2 (G2) acts as a trigger (shown with red in figure). c) Distribution of miRNA1 (M1) changes. d) The change at expression (shown with light red because minor changes are occured on their expression values.) of common target effects changes of proportional distribution of miRNA2 (M2). Expression values are rounded to integers for simplicity](Fig1.jpg){#fig1}

[#FIX]: <> "caption needs more text, every color should be explained"
[answer]: <> "explained in caption."

@@

The repression efficiency of a miRNA on the individual target ($Eff_{gi}$) is calculated according to equation \eqref{eq:1}; where miRNA expression ($C_m$) in local neighborhood is distributed among targets using individual gene expression levels ($C_{gi}$) (#FIX isn't $C_g$ gene expressions in the group is equal to total of $C_{gi}$ in group? answer: its my false. now its corect.). For the genes targeted by multiple miRNAs, cooperative activity of miRNAs on a target gene, $R$, is calculated by summing repression activity of each miRNA (Equation \eqref{eq:2}).

@@

### Multifactorial calculations in miRNA:target network
Interactions between miRNAs and their targets can be affected from various factors. So, our model integrates multiple factors when calculating overall miRNA activity. We classified factors into two categories. Factors effecting binding determine interaction between miRNA and target and they alter amount of miRNA sequestered to target. Degradation efficiency factors assign amount of degradated target amount in miRNA:target pairs.(#FIX fix this sentence: answer:). In other words, binding factors exert their influence before or during binding, efficiency factors exert their influence after binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017] in miRNA:target interactions are described as factors effecting binding affinity. Efficiency factors determine how many of miRNA:target complexes will result in inhibition and binding region on target drastically effect miRNA degradation efficiency [@hausser_analysis_2013; @helwak_mapping_2013]. Both binding and efficiency factors are normalized to their maximum values and scaled to [0,1] interval. The normalized values of factors take into account to determine binding activity and miRNA efficiency on targets (\autoref{fig2}). Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:3}; $C_m$, miRNA expression in the group; $C_{gi}$, individual gene expression; gi, individual gene (\autoref{fig2}c). 
\begin{equation}
Eff_{gi}= C_m \times E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}/(\sum_{1}^{i} E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}) \tag{3}\label{eq:3}
\end{equation}

@@

![Calculations to determine of miRNA binding and repression efficiency. $G$, Gene; $M$, miRNA; $STE$, seed type effect; $RE$, Region Effect; $E$, Energy; $STE^\prime$, normalized values of seed type efficiency coefficient; $RE^\prime$, normalized values of region efficiency coefficient; $E^\prime$, normalized values of energy coefficient.](fig2.jpg){#fig2}

\iffalse
%% (#TODO the equation is not considering binding factors, right?. We should explain the calculation made in  simulate function  here, in simple sentences. the function looks like this: 
%%    input_graph%>%
%%      tidygraph::activate(edges)%>%
%%      group_by(to)%>%
%%      mutate(mirna_count_per_comp = mirna_count_current*comp_count_current*afff_factor/sum(comp_count_current*afff_factor), mirna_count_per_comp = ifelse(is.na(mirna_count_per_comp), 0, mirna_count_per_comp))%>%
%%      ungroup()%>%
%%      mutate(effect_pre = effect_current, effect_current = mirna_count_per_comp*degg_factor)%>%
%%      group_by(from)%>%
%%      mutate(comp_count_pre = ifelse(comp_count_current< 0, 1, comp_count_current), comp_count_current = comp_count_pre - (sum(effect_current)-sum(effect_pre)), comp_count_current = ifelse(comp_count_current< 0, 1, comp_count_current))%>%
%%      ungroup()%>%
%%      mutate(comp_count_list = pmap(list(comp_count_list, comp_count_current), c), effect_list = pmap(list(effect_list, effect_current), c), mirna_count_list = pmap(list(mirna_count_list, mirna_count_current), c))
%% )
\fi

<answer> I shown in figure and explained at following. I didn't fully understand whether it is enough or not.

After miRNA binds to its target, but might not repress to bound target. The functionality of bound miRNA on target depends on efficiency factors like region that is binding sequence of miRNA on its target. Exact repression efficiency of miRNA is calculated according to equation \eqref{eq:4} (\autoref{fig2}d); $RE^\prime_{gi}$, normalized values of region efficiency coefficient between miRNA and gene. The cooperative repression activity of miRNAs to their common targets is figured out as shown in \autoref{fig2}e.

@@

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single  patient are retrieved from The Cancer Genome Atlas [TCGA](https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga)(#TODO). High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from ... (#FIX name of database and its citation) answer->suplementary datasets of [@helwak_mapping_2013; @moore_mirnatarget_2015]. Detailed description of network construction and its code is available in Supplementary data (#TODO burada link nasıl olacak, word dosyası veya PDF dosyası ismi mi vermemiz mi  gerekiyor, yoksa sadece Sup Data denilmesi yeterli mi? ilgili dosya: TCGA_E9-A1N5_article.Rmd). ABCC1 gene, over-expressed in tumor tissue, was selected as trigger for simulation of integrated dataset. After simulation of network, we have compared simulation results and tumor tissue expression levels.

@@

## Results and Conclusions
### Networks with single factor
We have developed a network-based approach to assess effects of expression level changes in competitive ceRNA regulation. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches [@arvey_target_2010; @denzler_assessing_2014]. Our approach can effortlessly calculate effects of expression changes when abundance levels of miRNAs and targets is only available factor. In sample network given in Figure (\autoref{fig1}), after an increase in expression level of a gene (G2), expression values of other genes also changed due to redistribution of miRNA among its targets. Previous studies have shown that if a gene abundance increases in ceRNA system, expression levels of genes targeted by shared miRNA are also affected [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that expression levels of primary neighborhoods which do not interact with another miRNA of the trigger gene change in relation to their expression. However, expression level change of common target (G4) in network has been observed different from G1 that has the same expression value with G4 target at initial conditions because more than one miRNA repress this target (\autoref{fig1}d) (#FIX last sentence needs some clarification, after "but cooperative...").

Genes targeted by multiple miRNAs act as a trigger for adjacent local neighborhood of targeting miRNAs, causing changes in expression levels of genes outside the local neighborhood of original trigger gene. Therefore, primary expression level change in gene (G2) causes changes in other group of genes (G5 and G6) even though original trigger gene (G2) and genes in other group are not targeted by common miRNA. In addition, as shown by ceRNA hypothesis model of Ala et al., after the increase of gene expression level of G2, the miRNA that is found in the same group (M1) tended to be less repressive on its remaining targets (G1, G3 and G4). It's important to note that the changes in gene expression levels will have more pronounced effect if miRNA:target ratio is high, i.e., more miRNA available per target, which was reported in previous findings [@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014]. [][#FIX]

[#FIX]: <> (following 4 sentences can be deleted, not so clear. OR we can summarize it one-two sentences. When the simulation was run in sample dataset that includes lower target abundance, the first response of primary neighborhood of the same group was determined as similar with in system that has higher total target abundance. In overall state in the end of simulation, we have established that all regulations more prominent when compared with low miRNA:target ratio dataset.)

@@

### Competing Endogenous RNA:Target networks based on interaction factors

[#FIX]: <> "title should be more informative"-> answer.

Earlier studies reported that miRNA regulatory interactions are affected by different parameters. For example, Xu et al. have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomics dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical for miRNA efficiency. In addition, binding energy between miRNA and targets is a significant determinant for miRNA efficiency and it has been reported that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. Similarly, another study revealed that affinity is correlated with seed pairing of miRNA:target pairs and suggested that affinity is correlated with number of canonical seed base pairing [@bosson_endogenous_2014]. Therefore, we have integrated aforementioned interaction parameters that could be useful for predicting miRNA repression activity more accurately. The sample dataset used in \autoref{fig1} is recalculated with additional factors \autoref{tab:one} in effect with same trigger, two fold increase in Gene2 (\autoref{fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in \autoref{fig3}a. Although the miRNA:target expression ratios in steady-state were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed.

[#FIX2]: <> "we should compare and contrast Figure 1a and figure 3a, the reader should see how affinity effects distrbution G4:M2, and how efficiency is effected without change in affnity M1:G3"
[answer2]: <> "I explained the mechanism, but i did not understand  how efficiency is effected without change in affnity M1:G3"
It is considered that entire miRNAs in the system affect targets according to target:total target ratio \autoref{fig1}a, when interaction factors did not take into account. However, in presence of interaction factors in miRNA:target interaction network (shown in \autoref{tab:one}), miRNAs are distributed according to target:total target ratio and affinity factors in first like shown in \autoref{fig2}c. After affinity mediated propotional distribution of miRNA expression, degradation factor is considered to specify count of repressive miRNA in pairs because entire bound miRNA target pairs might not be resulted with degradation (It is figured out shown in \autoref{fig2}d). For this reason, miRNA1 (M1) has more weak repressive effect (shown as edge variables) on target Gene2 \autoref{fig3}a in comparison with miRNA1 (M1):Gene2 (G2) interaction in \autoref{fig1}a. On the other hand, it is possible that taking into accont of these interaction factors could cause increasing miRNA repression activity like in miRNA2:Gene4 interaction \autoref{fig3}a. When expression of Gene2 (G2) increased, expression values of all genes also changed differentially because of contribution of efficiency factors (\autoref{fig3} b,c,d). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression.

@@

In regulation of miRNA target sample system (\autoref{fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA on the targets are weak after the increase of miRNA target G2. When model is triggered with two fold increase in expression level of common target (G4), the changes of other gene expressions have observed more prominent [][#TODO]. Furthermore, change in expression level of target gene that has strong miRNA repression efficiency resulted in evident perturbation in network. On the other hand, it was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors. Since perturbation efficiency of each gene is different, we have developed a function which screens each gene in the network for their perturbation efficiencies. **(see details in Supplementary File(fig1_2app.Rmd)**. On the other hand, when we applied the method on the minimal dataset, Gene4 has been found to be the most efficient element in terms of number of perturbed elements and miRNA2 (M2) has been found to be causing the highest mean in expression changes. [][#TODO2]

[#FIX]: <> "didnt we perform upregulation of Gene2, why did we say Gene4 here?" 
[answer]: <> "Because, in fact, We want to show that triggering element is significant in the same dataset, in the same conditions. Maybe I must add a conjuction here."

We used our model to simulate a real dataset which contains thousands of genes and hundreds of miRNAs. Our  model can successfully simulate perturbations in such large network despite complex behaviors and struggle to reach steady-state **(see details in Supplementary File(TCGA_E9-A1N5_article.Rmd)**. Simulations show that change in expression level of single gene has potential to effect whole network, perturbing almost all nodes. These observations are in accordance with competing endogenous RNA hypothesis where genes targeted with many common miRNAs subsequently transmit perturbation to neighboring groups.

Network based approaches for analyzing miRNA:target interactions have been developed in earlier studies. An initial  attempt demonstrated the ceRNA crosstalk in a network-like minimal interaction structure with concentrations of ceRNA and miRNAs [@figliuzzi_micrornas_2013]. Next, a network based kinetic model integrating miRNA and target rates of transcription, degradation, binding and unbinding was developed [@nitzan_interactions_2014] using high throughput experimental dataset about miR-92a depletion [@helwak_mapping_2013]. It was demonstrated that distant ceRNAs can interact with each other via indirect links, and the interactions are effected depending on distance between ceRNAs or topological features of network [@nitzan_interactions_2014]. More recently, an approach to detect ceRNA interaction by using the miRNA expression, gene expression and common miRNAs between gene targets was developed [@markus_list_sponge_2017] which was effective in analyzing genes through miRNAs. Based on the observations that a miRNA can exhibit strong functionality to a target but may not against an other, the authors have concluded that existing miRNA based approach may not be suitable for understanding regulations of ceRNA interactions.

[#Add]More recently, Silveira et al. have developed miRmapper package for using with R programme [@da2018mirmapper]. Researchers have utilized the adjacency matrix to associate miRNAs using differentially expressed genes and found similarities of miRNAs with this approach. Authors who utilized the approach in dataset of bladder cancer cell line have identified significant miRNAs in network, and miRNAs which work synergistically.

[#Add] Today, there are many of tools that provide predicted or experimentally validated miRNA:target interaction dataset. The datasets that contain weak, proteomic analysis like western blot, and strong, crosslinking and immunoprecipitation CLiP, experimentally evidence are more reliable sources than prediction datasets. While in experimentally weak supported datasets it is not known that indirect or direct target of miRNAs in cell, high-throughput methods ensure handling of AGO (Argounate Protein) interacted miRNAs. For this reason we have preferred to utilize high-throughput datasets in our application. Especially, we collected datasets that depend on chimeric [@helwak_mapping_2013; @moore_mirnatarget_2015] reading of miRNA: target pairs from among these datasets because they contain exact complementary seed regions.

[#Add] Although these experimental sources include exact binding sites, they do not provides which miRNA is functional or not [@liu2019prediction]. We have used the features of bound miRNA:target pairs and scaled these parameters inside groups that is consisted according to, ceRNA (Competing Endogenous RNA), genes targeted by the same miRNA. In scaling process, we specified the interaction factors as numeric values in accordance with previous studies. Energy values in miRNA:target pairs are represented by high-throughput studies [@helwak_mapping_2013; @moore_mirnatarget_2015] which are utilized in this study. On the other hand, we have specified the other interaction factors, seed type and location of binding region on the target, as numeric values based on the previous studies.[@grimson_microrna_2007] have compared the seed types' effect on target repression with few miRNA had canonical seed pairing in their study. Additionally, [@bartel_micrornas:_2009] and [@betel2010comprehensive] have studied on functional and non-functional seed interactions. Based on results of these studies we have arranged seed types of miRNA:target interactions as numeric values. We also have redefined location of binding region on the target as numeric values, based on studies of [@hausser_analysis_2013] and [@helwak_mapping_2013]. With this process, we have handled this entegrated dataset in context of competitor behaviours and functionality of interactions.

>>>>

Competing endogenous RNA (ceRNA) regulations and crosstalk between various types of non-coding RNA in human is remarkable in means of miRNA regulation. Many studies have pointed out that an alteration in miRNA:target interaction can result in unexpected changes due to indirect and complex interactions. In this paper, we defined a new network-based model that handles miRNA:ceRNA interactions with expression values. Our model is able to handle miRNA interaction factors such as seed type, binding energy, if provided. Our approach is able to reveal that a perturbation in an element of network affects whole competing elements differently and cooperative efficiencies of miRNAs on common targets could be calculated. Our findings emphasized importance of miRNA:target ratios being crucial, as reported by previous studies. We have showed that the competing elements which have the same or close expression values may not be affected equally from the perturbation because of repression functionality depended on interaction factors of miRNA target pairs. We applied the model to real sample consisting of breast cancer gene and miRNA expression dataset and experimental miRNA:target interaction dataset all generated via high throughput sequencing methods. A gene over-expressed in tumor tissue, namely *ABCC1*, is used as perturbing element. We have observed that change in expression level of single gene in miRNA:target network is sufficient to perturb regulations in whole network, due to unforeseen and unpredicted regulation which are only visible when considered in network context. Therefore, this model helps unveiling the crosstalk between elements in miRNA:target network where abundance of target and sponge effect are taken into account. The model is scalable and can be plugged in with emerging miRNA effectors such as circRNAs. The model is available as R package [ceRNAnetsim](https://github.com/selcenari/ceRNAnetsim).

@@

MicroRNAs (miRNAs) are a family of short non-coding RNAs which are key regulator of gene expression through various post-transcriptional mechanisms. Although the mechanisms by which miRNA represses are not fully understood, miRNAs predominantly repress their targets. Repressive activities of miRNAs vary depending on many factors that are significant to miRNA:target interactions. These factors include miRNA:target binding energy, binding location in target sequence, base pairing types between miRNA and target, abundance of miRNAs and targets [@grimson_microrna_2007]. Binding energies of miRNA:target complexes vary based on nucleotide context and determine folding stability of complex [@cao_predicting_2012]. It has been demonstrated that the binding energy between miRNA and target indicates stability or affinity of complex [@helwak_mapping_2013] and does not directly determine repressive activity of miRNA [@cao_predicting_2012]. Early studies have argued that 2-8 nt sequence, seed, located in miRNA 5'end bind to specific sequence located in 3'UTR of its target [@bartel_micrornas_2004; @lewis_conserved_2005]. In recent studies, it has been shown that miRNAs can interact with targets via sequences located in regions such as 5’UTR or CDS [@hausser_analysis_2013; @helwak_mapping_2013; @moore_mirnatarget_2015]. These studies also showed that  binding location either indicate functionality of miRNA:target interaction or affect abundance of targets. It has been shown that miRNAs exhibit repressive activity via, 6-8 nt long sequence that is perfectly complementary with targets, seed region at the 5’ end of miRNAs [@bartel_micrornas:_2009; @grimson_microrna_2007]. On the other hand, some researchers have reported that seed sequence of miRNA can have mismatches or bulged/wobble nucleotides [@chi2012alternative] and may locate in region other than 5'end of miRNAs [@hafner_transcriptome-wide_2010; @helwak_mapping_2013]. On top of all these factors, abundance of miRNAs and targets and miRNA:target ratio in cells predominantly affect efficiency of miRNA:target interaction [@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014].

@@

![Schematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: triangles) repress targets (G: circles) according to proportion of target expression level. b) Two fold increase in transcript level of Gene2 (G2) acts as a trigger (shown in red). c) Distribution of miRNA1 (M1) changes. d) The change at expression of common target effects changes of proportional distribution of miRNA2 (M2). Expression values are rounded to integers for simplicity. Gray values on edges indicate initial distribution, red values indicate increase and blue values indicate decrease in distribution. Shades of circles indicate different levels of increase.](Fig1.jpg){#fig1}

@@

The repression efficiency of a miRNA on the individual target ($Eff_{gi}$) is calculated according to equation \eqref{eq:1}; where miRNA expression ($C_m$) in local neighborhood is distributed among targets using individual gene expression levels ($C_{gi}$). For the genes targeted by multiple miRNAs, cooperative activity of miRNAs on a target gene, $R$, is calculated by summing repression activity of each miRNA (Equation \eqref{eq:2}).

@@

### Multifactorial calculations in miRNA:target network
Interactions between miRNAs and their targets can be affected from various factors. So, our model integrates multiple factors when calculating overall miRNA activity. We classified factors into two categories. Binding factors determine interaction between miRNA and target and they alter amount of miRNA sequestered to target. Efficiency factors dictate degradation efficieny of miRNA on its target and define amount of target to be degraded. In other words, binding factors exert their influence before or during binding, efficiency factors exert their influence after binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017] in miRNA:target interactions are described as factors effecting binding affinity. Also, binding region (e.g, 5'UTR, 3'UTR, CDS) was shown to drastically effect miRNA degradation efficiency [@hausser_analysis_2013; @helwak_mapping_2013]. Since binding or degradation efficiency from literature have various and broad ranges, both binding and efficiency factors are normalized to their maximum values and scaled to [0,1] interval. The normalized values of factors are used to determine binding activity and miRNA efficiency on targets (\autoref{fig2}). Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:3}; $C_m$, miRNA expression in the group; $C_{gi}$, individual gene expression; gi, individual gene (\autoref{fig2}c). 
\begin{equation}
Eff_{gi}= C_m \times E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}/(\sum_{1}^{i} E^\prime_{gi} \times STE^\prime_{gi} \times C_{gi}) \tag{3}\label{eq:3}
\end{equation}

@@

![Calculations to determine of miRNA binding and repression efficiency. $G$, Gene; $M$, miRNA; $STE$, seed type effect; $RE$, Region Effect; $E$, Energy; $STE^\prime$, normalized values of seed type efficiency coefficient; $RE^\prime$, normalized values of region efficiency coefficient; $E^\prime$, normalized values of energy coefficient. Numbering on edges match the pair order in \autoref{tab:one}](fig2.jpg){#fig2}

Not all miRNA:target binding events result in degradation of target. The degradation of target by bound miRNA depends on efficiency factors such as binding region. Exact repression efficiency of miRNA is calculated according to equation \eqref{eq:4} (\autoref{fig2}d); $RE^\prime_{gi}$, normalized values of region efficiency coefficient between miRNA and gene. The cooperative repression activity of miRNAs to their common targets is figured out as shown in \autoref{fig2}e.

@@

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single patient are retrieved from TCGA Research Network: https://www.cancer.gov/tcga. High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from published datasets [@helwak_mapping_2013; @moore_mirnatarget_2015]. Detailed description of network construction and its code is available in Supplementary Materials and Methods. ABCC1 gene, over-expressed in tumor tissue, was selected as trigger for simulation of integrated dataset. After simulation of network, we have compared simulation results and tumor tissue expression levels.

@@

## Results and Conclusions
### Networks with single factor
We have developed a network-based approach to assess effects of expression level changes in competitive ceRNA regulation. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches [@arvey_target_2010; @denzler_assessing_2014]. Our approach can effortlessly calculate effects of expression changes when abundance levels of miRNAs and targets is only available factor. In sample network given in Figure (\autoref{fig1}), after an increase in expression level of a gene (G2), expression values of other genes also changed due to redistribution of miRNA among its targets. Previous studies have shown that if a gene abundance increases in ceRNA system, expression levels of genes targeted by shared miRNA are also affected [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that expression levels of primary neighborhoods which do not interact with another miRNA of the trigger gene change in relation to their expression. However, changes in expression levels of G4 and G1 differ although they have same initial conditions. This is due to fact that G4 is targeted by two miRNAs, M1 and M2 (\autoref{fig1}d).

Genes targeted by multiple miRNAs act as a trigger for adjacent local neighborhood of targeting miRNAs, causing changes in expression levels of genes outside the local neighborhood of original trigger gene. Therefore, primary expression level change in gene (G2) causes changes in other group of genes (G5 and G6) even though original trigger gene (G2) and genes in other group are not targeted by common miRNA. In addition, as shown by ceRNA hypothesis model of Ala et al., after the increase of gene expression level of G2, the miRNA that is found in the same group (M1) tended to be less repressive on its remaining targets (G1, G3 and G4). It's important to note that the changes in gene expression levels will have more pronounced effect if miRNA:target ratio is high, i.e., more miRNA available per target, which was reported in previous findings [@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014].

@@

### ceRNA:Target networks based on interaction factors
Earlier studies reported that miRNA regulatory interactions are affected by different parameters. For example, Xu et al. have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomics dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical for miRNA efficiency. In addition, binding energy between miRNA and targets is a significant determinant for miRNA efficiency and it has been reported that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. Similarly, another study revealed that affinity is correlated with seed pairing of miRNA:target pairs and suggested that affinity is correlated with number of canonical seed base pairing [@bosson_endogenous_2014]. Therefore, we have integrated aforementioned interaction parameters that could be useful for predicting miRNA repression activity more accurately. The sample dataset used in \autoref{fig1} is recalculated with additional factors \autoref{tab:one} in effect with same trigger, two fold increase in Gene2 (\autoref{fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in \autoref{fig3}a. Although the miRNA:target expression ratios in steady-state were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed.

[#TODO]: <> "figure1, 2 ve 3 uzerinden farkli dagilim ve degradasyon açıklaması buraya. G2 uzerinden olmasa iyi olur"

It is considered that entire miRNAs in the system affect targets according to target:total target ratio \autoref{fig1}a, when interaction factors did not take into account. However, in presence of interaction factors in miRNA:target interaction network (shown in \autoref{tab:one}), miRNAs are distributed according to target:total target ratio and affinity factors in first like shown in \autoref{fig2}c. After affinity mediated propotional distribution of miRNA expression, degradation factor is considered to specify count of repressive miRNA in pairs because entire bound miRNA target pairs might not be resulted with degradation (It is figured out shown in \autoref{fig2}d). For this reason, miRNA1 (M1) has more weak repressive effect (shown as edge variables) on target Gene2 \autoref{fig3}a in comparison with miRNA1 (M1):Gene2 (G2) interaction in \autoref{fig1}a. On the other hand, it is possible that taking into accont of these interaction factors could cause increasing miRNA repression activity like in miRNA2:Gene4 interaction \autoref{fig3}a. When expression of Gene2 (G2) increased, expression values of all genes also changed differentially because of contribution of efficiency factors (\autoref{fig3} b,c,d). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression.

@@

In regulation of miRNA target sample system (\autoref{fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA on the targets are weak after the increase of miRNA target G2. When model is triggered with two fold increase in expression level of common target (G4), more prominent changes were observed in gene expression levels in network compared to changes caused by G2 increase. Furthermore, change in expression level of target gene that has strong miRNA repression efficiency resulted in evident perturbation in network. On the other hand, it was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors. Since perturbation efficiency of each gene is different, we have developed a function which screens each gene in the network for their perturbation efficiencies. **(see details in Supplementary File(fig1_2app.Rmd)**. On the other hand, when we applied the method on the minimal dataset, Gene4 has been found to be the most efficient element in terms of number of perturbed elements and miRNA2 (M2) has been found to be causing the highest mean in expression changes.

### Simulation of large and realistic network

Apart from tools predicting miRNA:target pairs, there are databases that provide  experimentally validated miRNA:target interaction datasets. The reliability in these datasets range from weak (e.g, western blot) to strong (e.g, CLiP, CLASH). High-throughput methods ensure handling of AGO (Argounate Protein) interacted miRNAs (CLiP) and their targets as well (CLEAR-CLiP, CLASH). We constructed a large ceRNA network utilizing high-throughput datasets containing chimeric reading of miRNA:target pairs [@helwak_mapping_2013; @moore_mirnatarget_2015] which expose exact complementary seed regions.

Although these experimental sources include exact binding sites, they do not imply that each miRNA binding event is functional [@liu2019prediction]. Seed type, energy, binding region affects functionality of miRNA drastically. In order to construct realistic network, we incorporated such factors in the network. Energy  values in miRNA:target pairs represented by high-throughput studies [@helwak_mapping_2013; @moore_mirnatarget_2015], seed type [@grimson_microrna_2007; @bartel_micrornas:_2009; @betel2010comprehensive] and binding region effect [@hausser_analysis_2013; @helwak_mapping_2013] data were collected and integrated  into network.

The network was finalized by integrating gene and miRNA expression levels from TCGA determined by RNA-Seq. Our  model can successfully simulate perturbations in such large network despite complex behaviors and struggle to reach steady-state **(see details in Supplementary File(TCGA_E9-A1N5_article.Rmd)**. Simulations show that change in expression level of single gene has potential to effect whole network, perturbing almost all nodes. These observations are in accordance with competing endogenous RNA hypothesis where genes targeted with many common miRNAs subsequently transmit perturbation to neighboring groups.

[#TODO] We have observed that the results of simulation with ABCC1 gene, *Multidrug resistance-associated protein 1*, which is one of the most significant factors to develop resistance against chemotherapotic agents[@atalay2008role; @lu2015microrna; @atalay2006multidrug] is not convenient with tumor tissue expression values, when we compared the tumor gene expression dataset retrieved from TCGA and simulation results obtained via our approach. Simulation results were not expected to be consistent with tumor tissue expression values because large number of abnormal regulations were observed in these tissues. So, it might not said that a single gene is responsible for entire regulations in cancer cells. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Therefore, it would be more useful to test our approach with an ideal dataset that includes entire informations of gene expression values before and after the regulation of a known gene at existing tissue, and miRNA expression values of this tissue at initial conditions.

[#TODO]: <> "run perturbation screening on tcga set, find a gene then couple sentences about that gene"

## Discussion

Network based approaches for analyzing miRNA:target interactions have been developed in earlier studies. An initial  attempt demonstrated the ceRNA crosstalk in a network-like minimal interaction structure with concentrations of ceRNA and miRNAs [@figliuzzi_micrornas_2013]. Next, a network based kinetic model integrating miRNA and target rates of transcription, degradation, binding and unbinding was developed [@nitzan_interactions_2014] using high throughput experimental dataset about miR-92a depletion [@helwak_mapping_2013]. It was demonstrated that distant ceRNAs can interact with each other via indirect links, and the interactions are effected depending on distance between ceRNAs or topological features of network [@nitzan_interactions_2014]. More recently, an approach to detect ceRNA interaction by using the miRNA expression, gene expression and common miRNAs between gene targets was developed [@markus_list_sponge_2017] which was effective in analyzing genes through miRNAs. Based on the observations that a miRNA can exhibit strong functionality to a target but may not against an other, the authors have concluded that existing miRNA based approach may not be suitable for understanding regulations of ceRNA interactions. Additionally, miRmapper package [@da2018mirmapper] utilizes an adjacency matrix to associate miRNAs using differentially expressed genes and identifies significant miRNAs in network.

In our approach, we have not taken into account transcription, degradation or binding rates of elements in network. Because, although it is known as the miRNAs are highly stable, the transcription and degradation rates of miRNAs change depending on cellular conditions [@ruegger_microrna_2012]. However, due to lack of datasets including degradation and transcription rates of miRNAs at specific cellular conditions, such factors are neglected in our model. Additionally, other regulation parameters such as gene-gene interactions and transcription factors are ignored but the network structure is flexible and can integrate additional regulation elements as they become available.

## Conclusion
